---
title: 'MRIPSundries'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MRIPSundries}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align='center',
                      fig.width = 6, fig.height = 3)
```

<!-- 
Put a flag object in the top of the doc taht shows whether the files are available … and run the code only if they are.

To build lib: (1) devtools::document() (2) devtools::build_vignettes() (3) devtools::install()

All devtools functions accept a path as an argument, e.g. load_all("path/to/path/mypkg"). If you don't specify a path, devtools will look in the current working directory - this is recommended practice.

Frequent development tasks:

    load_all() simulates installing and reloading your package, loading R code in R/, compiled shared objects in src/ and data files in data/. During development you usually want to access all functions so load_all() ignores the package NAMESPACE. load_all() will automatically create a DESCRIPTION if needed.

    document() updates documentation, file collation and NAMESPACE.

    test() reloads your code, then runs all testthat tests.

Building and installing:

    install() reinstalls the package, detaches the currently loaded version then reloads the new version with library(). Reloading a package is not guaranteed to work: see the documentation to unload() for caveats.

    build() builds a package file from package sources. You can use it to build a binary version of your package.
end -->



## Purpose
Gather together MRIP analysis functions to make life easier. In general, this package provides functions to:

* load raw MRIP .csv files (trip, catch, size) and bind them together into a single R object and calculate appropriately weighted total catch, size composition and fishing effort;

* Parse out dates and put them in a format readable by R.

* load MRIP estimate files (catch, effort) and bind them together into a single R object, calculating aggregated means and PSEs as necessary

The package does not include functions to help with plotting but there are some examples in this vignette about how the data aggregation functions in this package can easily prep the data for plots.

## MRIP data
The MRIPSundries package depends entirely on access to MRIP .csv files. The intercept files are publicly available and can be found [here on the MRIP website](https://www.st.nmfs.noaa.gov/st1/recreational/MRIP_Survey_Data/) and the estimate files can be found [here](https://www.st.nmfs.noaa.gov/st1/recreational/MRIP_Estimate_Data/). Before using this package, download the data (all of it or whatever you need of it). The default file structure is fine, or if you want to re-organize the files that's fine too -- you just need to tell the package what the root directory is for the files and it ought to work.






## Combining .csv files
The .csv files hosted by MRIP are organized into trip, catch and size files, by wave -- in other words each wave in each year has it's own trip, catch and size .csv. To make things easier to work with, the ```bindMRIP()``` function takes the .csv files you are interested in and binds them together in a data frame.

For this vignette I assume that the files exist in a directory called mripDat which is a folder in the working directory.

Before doing anything else it would be useful to load the MRIPSundries library!

```{r message=FALSE, warning=FALSE}
library(MRIPSundries)
```

First let's check out what files we have:

```{r}
getwd()
fl <- list.files('vinData', full.names = TRUE)
print(head(fl))
```

So we have catch, size and trip mrip files from waves 4 & 5 in 2050-2052. I chose these years to avoid any confusion that the data in this vignette might be real data! If it's already 2052 and I've confused you, sorry. Before we can do much we first have to take these .csv files of each type and bind them together. To do this we can use the ```bindMRIP()``` function. First, just to show everything we have, let's load in some catch data and not do any subsetting. All we have to do is provide a directory (remember, this assumes that mripDat is a folder in our working directory) and the type of files we want to load.

```{r, message=FALSE}
# Bind together catch data
mDatC <- bindMRIP(dir = 'vinData',      # Directory holding the .csv files
                  type = 'catch',     # Type of .csv files to bind together
                  years = 2050:2052,  # Years to include
                  waves = 4:5)        # Waves to include
```

The bindMRIP function is useful, but note that reading in and binding together many large .csv files can take a while. Different types of data can be much faster to load in. If you keep re-running code, one solution is to run the ```bindMRIP()``` function once, then save the output as an rds object using ```saveRDS()```; after that you can load it back in using ```loadRDS()``` and you can skip ```bindMRIP()```.

### Checking out the data

We have standard tools to check out mDatC such as ```summary()``` and ```head()``` but a useful function from this package is ```get_dataStats()```

```{r}
get_dataStats(mDatC)
```

From here we can see that we have years from 2050-2052, two distinct species (if there were only one it would print the name), modes 3, 4, 5 and 7 and waves 4 and 5 (both of which are present in all three years).

Let's take a peek at the column names. Note that the actual MRIP data sets have more column names -- this is cut down for simulated data.

```{r}
names(mDatC)
```

Note that while the column names are the same as in the raw data files all the columns coming out of ```bind_mrip()``` are capitalized. This was an easy fix for some changes that had been made to the raw data standards over time and may be fixed in the future for consistency with the .csv files.

Sometimes exact definitions can be confusing to remember. We can use the ```printColDef()``` function to print out the definition of one or more columns.

```{r}
printColDef(colName = c('MODE_FX'))
```

The function will search through variable names and definitions from the trip, catch and size data bases. The definitions come from the MRIP website [here](https://www.st.nmfs.noaa.gov/st1/recreational/MRIP Survey Data/) under *MRIP survey variables.xlsx*. Note that we have to spell the <colName> arguments correctly correctly but we don't have to worry about case in this function.



## Data summaries
### Calculating summed or averaged catch / harvest

The ```get_catch()``` function calculates catch in numbers and weight. The arguments are

* **cdat**, the catch data set (typically built using bindMRIP() with the *type* argument ‘catch’;

* **group** which indicates the structure of the output (e.g., do you want catch by just mode? Mode and wave?); and

* **agFun** which is the aggregation function. This will typically be ```sum``` if calculating annual statistics (i.e., if you include YEAR in the *group* argument) but could be something like mean if you wanted to know the mean catch over a range of years (i.e., if you do not include YEAR in *group*).

The output is harvest in numbers (HN), releases in numbers (RN), harvest in kilograms (HKG) and harvest in pounds (HLB). Let’s find the annual catch for our data just by year

```{r}
catY <- get_catch(mDatC,                # data set from bindMRIP()
                  group = c('YEAR'),    # structure of output
                  common = 'BLUEFISH')  # species
knitr::kable(catY, digits=0)            # print a nice table
```

and now by year and mode (not all rows shown)

```{r}
catYM <- get_catch(mDatC,                         # Data from bindMRIP()
                   group = c('YEAR', 'MODE_FX'),  # columns to group by
                   common = 'BLUEFISH')           # species
knitr::kable(head(catYM), digits=0)               # print a nice table
```


### Calculating catch composition

The ```get_harvestSizeComp()``` function calculates the size composition of the harvest, appropriately weighted by WP_CATCH. The arguments are

* **sdat**: the size data frame, that typically comes from a call to ```bindMRIP()``` with argument *type* set to size;

* **group**: A vector of column names to structure output over (Year, wave, MODE_FX, etc.); and

* **units**: a string, either ‘in’ for inches or ‘cm’ for centimeters.

Let’s test the function, looking at the size composition over year and mode. Before we do this though we have to use ```bindMRIP()``` do develop a data frame from the size csv files:

```{r, message=FALSE}
# Bind together size data
mDatS <- bindMRIP(dir = 'vinData',     # Directory holding the .csv files
                  type = 'size')     # Type of file to bind together
```


Now we can get the size distribution in inches (only the first 15 rows printed here):

```{r}
sizYW <- get_harvestSizeComp(sdat = mDatS,               # Data set from bindMRIP()
                             group = c('YEAR', 'WAVE'),  # Structure of output
                             common = 'BLUEFISH',
                             unit = 'in')                # Desired units
knitr::kable(head(sizYW, 15), digits = 0)                # Print a nice table
```

### Calculating effort

The ```get_effort()``` function calculates fishing effort in number of trips, appropriately weighted by WP_INT. The arguments are

* **tdat**: the trip data frame, that typically comes from a call to bindMRIP() with argument type set to 'trip';

* **common**: Common name of the species of interest

* **group**: A vector of column names to structure output over (Year, wave, MODE_FX, etc.); and

* **type**: A vector of the types of trips to include in the effort calculation. Options are:
  * PRIM1_COMMON: primary target
  * PRIM2_COMMON: secondary target
  * A: type A catch recorded during trip
  * B1: type B1 catch recorded during trip
  * B2: type B2 catch recorded during trip

* **cdat**: the catch data frame, typically coming from a call to bindMRIP() with argument *type* set to 'catch' (argument is optional and only necessary if *type* includes 'A', 'B1' or 'B2';

Let’s test the function, and define fishing effort as all trips where bluefish were targeted as either the primary or secondary target. We will also subset the output to look only at Massachusetts. We already have a catch data set **mdatC** so we will only need to bind together the trip data. Note that we have to use the same year and wave parameters for both the catch and the trip data sets or the function will produce an error.

```{r, message = FALSE}
# Bind together trip data
mDatT <- bindMRIP(dir = 'vinData',      # Directory holding the .csv files
                  type = 'trip',      # Type of .csv files to bind together
                  years = 2050:2052,  # Years to include
                  waves = 4:5)        # Waves to include

eff <- get_effort(tdat = mDatT,                              # Trip data set
                  common = 'BLUEFISH',                       # species of interest
                  group = c('YEAR', 'ST'),                   # grouping over
                  type = c('PRIM1_COMMON', 'PRIM2_COMMON'))  # effort definition

knitr::kable(eff, digits = 0)
```

We can also partition the trip data over other groupings such as fishing mode (MODE_FX). In this example we specify effort as all trips where bluefish were targeted or landed (i.e., everything but *type* = 'B2'. Because we include landed catch we have to include the argument *cdat*. To avoid issues we have to specify a catch data set that includes only the species we are interested in (bluefish)

```{r, message = FALSE}
sp <- 'BLUEFISH'
mDatCBlue <- filter(mDatC, COMMON == 'BLUEFISH')

eff2 <- get_effort(tdat = mDatT,                            # Trip data set
                   cdat = mDatCBlue,                        # Catch data set
                   common = 'BLUEFISH',                     # species of interest
                   group = c('YEAR', 'ST', 'MODE_FX'),      # grouping over
                   type = c('PRIM1_COMMON', 'PRIM2_COMMON',
                           'A', 'B1')) %>%                  # effort definition
  filter(ST == 25 & YEAR == 2050)                           # Just worry about MA in 2050
  
knitr::kable(eff2, digits = 0)
```

## Adding dates

The trip, catch and size data frames all have month associated with each record. However, it could be useful to have finer-scale information on the distribution of records. ```The get_date()``` function parses the ID_CODE column to pull out the year/month/day information and then converts this to an object of class Date which is interpretable by R.

```{r}
d <- get_date(mDatC)
print(head(d))
```


## MRIP estimate files
The MRIP estimate files include both catch and effort data. The catch data include many of the same columns as the intercept data, just aggregated over groupings such as state, mode, etc. and, importantly, they include the variance. This allows for calculating PSEs over any grouping[^1] and the effort data include columns for estimated number of trips and the variances over the typical grouping levels.

[^1]: Note that (1) the variance of a sum equals the sum of the variances, making it easy to combine uncertainty at different levels of aggregation; and (2) PSE is the same as CV -- the coefficient of variation -- which is the standard deviation divided by the mean.

This vignette does not currently have dummy data for the MRIP estimates. However, implementation is much the same as for the ```bindMRIP()``` function above and will suffice for now. The function arguments are:

* **dir** the directory containing the MRIP estimate files
* **type** the desired type of estimates -- 'catch' or 'trip'
* **years** the years of interest (this governs which files are loaded in -- further groupings are available using the XXXXXX function).

```
bindMRIPEstimates(dir = 'estData',             # path to folder containing .csv files
                  type = 'catch',              # specifies catch data sets
                  years = 2050:2052)           # years to include
```

Aggregation of the data over different levels is through the XXXXXX function.

```
XXXXXX()
```

## Plotting

The package does not have specific plotting functions. However a couple of examples are given here using the ggplot2 package.

### Annual catch

Plot of annual catch (MA & RI combined) by mode (using the catYM data frame created above)

```{r}
catYM %>%        # References data set created above
  mutate(MODE_FX = as.character(MODE_FX)) %>%  # Treat MODE+FX as string
  ggplot(aes(x = YEAR, y = HLB,                # Assign x and y
             group = MODE_FX)) +               # Assign grouping variable
  geom_line(aes(color = MODE_FX),   # Adjust color by group
            size = 2) +             # Adjust line width
  geom_point(size = 2)              # Add points
```

Below is a plot of annual catch by wave, using a barplot. Here we have to re-run the ```get_catch()``` function to get output structured by wave. WAVE was changed to a character variable as an easy way to improve the apearance of the plot and legend.

```{r}
catYW <- get_catch(mDatC,   # Data frame created by bindMRIP()
                   group = c('YEAR', 'WAVE', 'ST'),     # Structure of output
                   common = 'BLUEFISH') %>%             # species name
  mutate(WAVE = as.character(WAVE))                     # Treat wave as string
ggplot(catYW, aes(x = YEAR, y=HLB,                      # Assign x and y
                  fill = WAVE)) +                       # Fill color is wave
  geom_bar(stat = 'identity') +     # Add barplot - identity treats values as-is
  facet_wrap(~ST)                   # Different panels by year
```

### Size composition


```{r}
ggplot(sizYW, aes(x = L_IN_BIN, y = HN)) +  # Assign x and y
  geom_bar(stat = 'identity') +             # Add barplot
  facet_wrap(~YEAR)                         # Panels are different years
```



### Cumulative catch

With date information we can do things like plot the cumulative distribution of catch during each year.

```{r}
mDatC %>%    # Data set created earlier
  add_column(Date = get_date(dat = mDatC)) %>%            # Add a date column
  filter(COMMON == 'BLUEFISH') %>%                        # Subset for bluefish
  arrange(Date) %>%                                       # Sort by date
  group_by(YEAR) %>%                                      # Year is grouping variable
  mutate(cumHarv = cumsum(LANDING * WP_CATCH / 1e6)) %>%  # Calculate cumulative harvest
  ggplot(., aes(x = Date, y = cumHarv)) +                 # Assign x and y
  facet_wrap(~YEAR,                                       # Panels should be years
             scales = 'free_x',                           # No common x labels
             nc = 2) +                                    # Two panel columns
  geom_line() +                                           # Include lines
  geom_point() +                                          # Include points
  ylab('Cumulative harvest (mil)')                        # Change y label
```



## 







